---
title: "The **BBMRIomics** package"
output: 
  html_document:
    toc: true
bibliography: BBMRIomics.bib
---

# Introduction to the **BBMRIomics** package #

The aim of the **BBMRIomics**-package is to make integrated analyses,
within the BIOS data collection or across BBMRI omics projects, as easy as
possible, using the `R`-programming language.

The **BBMRIomics** package is not a regular `R` package that lives on
`CRAN` or `Bioconductor`; it can only be installed on the surfSARA HPC
Virtual Machines. This is because datasets available on the surfSARA
infrastructure are linked during installation of the
package. Nevertheless, the use of the package is identical to any
other `R`-package.

## Loading the package ##

The package is loaded in the usual:

```{r}
library(BBMRIomics) ##load BBMRIomics package
```

On loading a *welcome message* is printed with some useful
information:
1. location of vignettes, 
2. wiki, 
3. who to contact with any issue's on the package or VM
4. where to post questions and importantly 
5. the version number of the package. 

We are continuing developing the package by adding new features or
datasets, e.g. depending on requests by users like you.

The **BBMRIomics** package is dependent on a few other packages,
especially the `Bioconductor` package
[**SummarizedExperiment**](https://www.bioconductor.org/packages/SummarizedExperiment/).

```{r}
sessionInfo() ##show which other packages are loading as well
```

## Getting help ## 

Both [wiki](http://wiki.bbmri.nl/wiki/BIOS_start-) and
[vignettes](http://bios-vm.bbmrirp3-lumc.surf-hosted.nl/BBMRIomics/)
contain useful information on the BBMRI omics data and usage of
surfSARA HPC Virtual Machines and the **BBMRIomics**
package. Furthermore, like any other `R`-package **BBMRIomics** has
buildin man-pages for the functions and datasets available.

```{r}
?getGenotypes
```

## Available datasets ##

```{r, eval=FALSE}
data(package="BBMRIomics")
```

```{r}
?metabolomics_RP3RP4_overlap
?methData_Betas_LLS_F2_cleaned
```

## Loading a dataset ##

Datasets can be loading using the `R`-function `data(<dataset_name>)`:

```{r}
data(methData_Betas_PAN_F2_cleaned)
ls()
```

The loaded object is called: `betas`. 

> NOTE: some datasets are contain Giga bytes of data so loading can take sometime!

# Using the datasets #



```{r}
betas
```

