---
title: "GRCh38 RNA Freeze2 unrelated datasets"
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
bibliography: BBMRIomics.bib
---

```{r, child="_setup_devel.Rmd"}
```

# Library preparation and alignment #

Detailed description of the RNASeq data generation and preprocessing
can be found in the supplementary material of the paper by Zhernakova
et al. [@Zhernakova2017]

# Feature quantification read counts #

Gene read counts are generated using
[featurecounts](http://bioinf.wehi.edu.au/featureCounts/).

Reference annotation used is from ensembl:
[release 91](ftp://ftp.ensembl.org/pub/release-91/gtf/homo_sapiens/Homo_sapiens.GRCh38.91.gtf.gz)

## Prepare gene annotation ##

For convenience we create a `TxDb`-package. For more information on
using a `TxDb`-package see
[GenomicFeatures](http://bioconductor.org/packages/release/bioc/vignettes/GenomicFeatures/inst/doc/GenomicFeatures.pdf).

*** Note this requires sudo rights to install and use sudo -i to load /etc/profile with global R_SITE_LIB =/opt/... ***
```{r, txdb, eval=FALSE}
library(GenomicFeatures)
library(rtracklayer)
gtf <- "ftp://ftp.ensembl.org/pub/release-91/gtf/homo_sapiens/Homo_sapiens.GRCh38.91.gtf.gz"
txdb <- makeTxDbFromGFF(gtf, organism="Homo sapiens")
##Hack by James MacDonald (https://support.bioconductor.org/p/89855/)
con <- dbconn(txdb)
## insert a Resource URL entry in the metadata table,
## pointing to the ftp site where the data were procured
DBI::dbGetQuery(con, "INSERT INTO metadata VALUES ('Resource URL', 'ftp://ftp.ensembl.org/pub/release-91/gtf/homo_sapiens/Homo_sapiens.GRCh38.91.gtf.gz');")
makeTxDbPackage(txdb,
                version = "0.0.1",
                maintainer = "D. Cats <d.cats@lumc.nl>",
                author = "d. cats",
                destDir= ".",
                license = "Artistic-2.0",
                pkgname = "TxDb.Hsapiens.Ensembl.v91")
library(devtools)
install("./TxDb.Hsapiens.Ensembl.v91")
```

Now we can use the Ensembl v71 annoation like any other `TxDb`-package.

```{r, txdbexample, eval=FALSE}
library(TxDb.Hsapiens.Ensembl.v91)
(txdb <- TxDb.Hsapiens.Ensembl.v91)
genes(txdb)
```

## Prepare sample metadata ##
Metadata for the samples is retrieved from the metadatabase. This is done only
for the samples which were part of maximum set of unrelated individuals from 
Freeze2. These are the same samples as were uploaded to EGA 
(Fastq files: [EGAD00001003785](https://ega-archive.org/datasets/EGAD00001003785)).

```{r, load_BBMRIomics, eval=FALSE}
library(BBMRIomics)
freeze2_unrelated <- runQuery(paste0(
    "SELECT uuid, biobank as biobank_id, bios_id AS ids, gonl_id, ",
        "rna_run.rna_run_id AS run_id, age AS sampling_age, ", 
        "COALESCE(rna_sampling_date, dna_bloodsampling_date, ",
            "cellcount_bloodsampling_date, lipids_bloodsampling_date, ",
            "crp_bloodSampling_date) AS sampling_date, ",
        "COALESCE(rna_sampling_time, dna_bloodsampling_time, ",
            "lipids_bloodsampling_time, crp_bloodsampling_time, ",
            "cellcount_bloodsampling_time) AS sampling_time, ",
        "sex, birth_year, weight, height, anthropometry_age, smoking, smoking_age, ",
        "ascertainment_criterion, totchol, triglycerides, hdlchol, ldlchol, ",
        "ldlcholmethod, hscrp, lipidmed, lipidsmed_age, lipids_bloodsampling_fasting, ",
        "rna_source, rna_rin, rna_a260a280ratio, rna_extraction_method, ",
        "rna_extraction_date, dna_source, dna_quantificationmethod, dna_a260a280ratio, ",
        "dna_extraction_method, dna_extraction_date, gwas_chip, ",
        "gwas_datageneration_date, baso, baso_perc, eos, eos_perc, granulocyte, ",
        "granulocyte_perc, luc, luc_perc, lymph, lymph_perc, mono, mono_perc, neut, ",
        "neut_perc, hct, hgb, mch, mchc, mcv, mpv, plt, rbc, rdw, wbc, ",
        "flowcell_num, machine, run_num, insert_size, rna_id ",
    "FROM rna_freeze ",
    "LEFT JOIN rna_run ON rna_freeze.rna_run_id = rna_run.rna_run_id ",
    "LEFT JOIN rna_sample ON rna_run.rna_sample_id = rna_sample.rna_sample_id ",
    "LEFT JOIN visit ON rna_sample.visit_id = visit.visit_id ",
    "LEFT JOIN person ON visit.person_id = person.person_id ",
    "LEFT JOIN gwas on person.person_id = gwas.person_id ",
    "LEFT JOIN dna_sample on visit.visit_id = dna_sample.visit_id ",
    "WHERE set = 'unrelated' AND freeze_num = 2;"),
  usrpwd = RP3_MDB_USRPWD)
```


### Biobanks ###
Some of the biobank ids need to be relabeled to be conformant with the previous
datasets.

```{r, rename_biobanks, eval=FALSE}
freeze2_unrelated[freeze2_unrelated$biobank_id == "VUNTR", "biobank_id"] <- "NTR"
freeze2_unrelated[freeze2_unrelated$biobank_id == "LLS_PARTOFFS", "biobank_id"] <- "LLS"
freeze2_unrelated[freeze2_unrelated$biobank_id == "LIFELINES", "biobank_id"] <- "LL"
```

### Imputation IDs ###
Information about genotypes are still missing from the metadata of the samples.
Genotype imputation ids are retrieved from the meta database and added to the
sample metadata.

```{r, imputation_ids, eval=FALSE}
imputation <- getSQLview("getimputations")
hrc <- subset(imputation, imputation_reference == "HRC")
hrc <- hrc[, c("ids", "imputations_id")]
names(hrc) <- c("ids", "imputation_id_hrc")
hrcv1.1 <- subset(imputation, imputation_reference == "HRCv1.1")
hrcv1.1 <- hrcv1.1[, c("ids", "imputations_id")]
names(hrcv1.1) <- c("ids", "imputation_id_hrcv1.1")
freeze2_unrelated <- merge(freeze2_unrelated, hrc[, c("ids", "imputation_id_hrc")],
                           by="ids", all.x=TRUE)
freeze2_unrelated <- merge(freeze2_unrelated, hrcv1.1[, c("ids", "imputation_id_hrcv1.1")], 
                           by="ids", all.x=TRUE)
```

### Normalize inconsistent phenotype values ###
There are a number of fields which contain inconsistent values. The potentially
affected fields are checked and adjusted accordingly.

```{r, normalize_phenotypes, eval=FALSE}
table(freeze2_unrelated$biobank)
table(freeze2_unrelated$sex)
table(freeze2_unrelated$smoking)
table(freeze2_unrelated$ascertainment_criterion) #needs some normalization
table(freeze2_unrelated$ldlcholmethod)
table(freeze2_unrelated$lipidmed)
table(freeze2_unrelated$lipids_bloodsampling_fasting)
table(freeze2_unrelated$rna_source)
table(freeze2_unrelated$rna_extraction_method) # needs some normalization
table(freeze2_unrelated$dna_source)
table(freeze2_unrelated$dna_quantificationmethod) # needs some normalization
table(freeze2_unrelated$dna_extraction_method) # needs some normalization
table(freeze2_unrelated$gwas_chip) # needs some normalization

## ascertainment_criterion
freeze2_unrelated$ascertainment_criterion <- gsub("^GoNL$", "GONL_subject", 
                                                  freeze2_unrelated$ascertainment_criterion)
freeze2_unrelated$ascertainment_criterion <- gsub("^Complete genomics sequencing$",
                                                  "CG_subject", 
                                                  freeze2_unrelated$ascertainment_criterion)
freeze2_unrelated$ascertainment_criterion <- gsub("^GoNL / Complete genomics sequencing",
                                                  "GoNL/CG_subject", 
                                                  freeze2_unrelated$ascertainment_criterion)
table(freeze2_unrelated$ascertainment_criterion)

## rna_extraction_method
freeze2_unrelated$rna_extraction_method[!is.na(freeze2_unrelated$rna_extraction_method)] <- 
    "PAXgene Blood RNA Kit (Qiagen)"
table(freeze2_unrelated$rna_extraction_method)

## dna_quantificationmethod
freeze2_unrelated$dna_quantificationmethod <- gsub("nanodrop|nano", "Nanodrop", 
                                                   freeze2_unrelated$dna_quantificationmethod)
freeze2_unrelated$dna_quantificationmethod <- gsub(
    "spectrofotometer|Spectofotometer|spectofotometer", "Spectrofotometer", 
    freeze2_unrelated$dna_quantificationmethod)
table(freeze2_unrelated$dna_quantificationmethod)

## dna_extraction_method
freeze2_unrelated$dna_extraction_method <- gsub("salting out|Saltingout", "Salting out", 
                                                freeze2_unrelated$dna_extraction_method)
freeze2_unrelated$dna_extraction_method <- gsub("QIAamp DNA minikit", "DNA Mini Kit (Qiaamp)", 
                                                freeze2_unrelated$dna_extraction_method)
table(freeze2_unrelated$dna_extraction_method)

## gwas_chip
freeze2_unrelated$gwas_chip <- gsub(" *, *", "/", freeze2_unrelated$gwas_chip)
freeze2_unrelated$gwas_chip <- gsub("Illumina human omni express", "Illumina OmniExpress", 
                                    freeze2_unrelated$gwas_chip)
freeze2_unrelated$gwas_chip <- gsub("Illumina660", "Illumina Human660-Quad", 
                                    freeze2_unrelated$gwas_chip)
freeze2_unrelated$gwas_chip <- gsub("OverlappingSNPsfrom", "Overlapping SNPs from ",
                                    freeze2_unrelated$gwas_chip)
freeze2_unrelated$gwas_chip <- gsub("GONLSequenceData", "GONL", freeze2_unrelated$gwas_chip)
table(freeze2_unrelated$gwas_chip)
```

### Fill in missing LDLchol and LDLcholMethod ###

For some samples the LDLchol values are missing, but can be inferred from the
totChol, HDLchol and TriGlycerides using the 
[Friedewald estimation](http://www.gpnotebook.co.uk/simplepage.cfm?ID=x20030114211535665170).

```{r, LDLchol, eval=FALSE}
table(is.na(freeze2_unrelated$ldlchol))

##Calculate LDLchol, for those missing, from Tot, HDL and TriGlycerides
##http://www.gpnotebook.co.uk/simplepage.cfm?ID=x20030114211535665170    
LDLchol <- freeze2_unrelated$totchol - freeze2_unrelated$hdlchol - (freeze2_unrelated$triglycerides/2.2)    
freeze2_unrelated$ldlcholmethod[is.na(freeze2_unrelated$ldlchol)] <- "Friedewald estimation"    
freeze2_unrelated$ldlchol[is.na(freeze2_unrelated$ldlchol)] <- LDLchol[is.na(freeze2_unrelated$ldlchol)]
## Friedewald estimation should not be used when triglycerides exceed 4.52
freeze2_unrelated$ldlcholmethod[freeze2_unrelated$triglycerides > 4.52 & 
                                    freeze2_unrelated$ldlcholmethod == "Friedewald estimation"] <- NA    
freeze2_unrelated$ldlchol[freeze2_unrelated$triglycerides > 4.52 & 
                              freeze2_unrelated$ldlcholmethod == "Friedewald estimation"] <- NA

table(is.na(freeze2_unrelated$ldlchol))
```

## Prepare count table ##
```{r, counts, eval=FALSE}
# prepare count table
library(BiocParallel)
register(MulticoreParam(32, log=TRUE))
files <- list.files(file.path(VM_BASE_DATA, "RNASeq/GRCh38/", "gene_read"), 
                    full.names=TRUE, pattern=".tsv$")

##select runs
files <- files[gsub("\\.tsv$", "", basename(files)) %in% freeze2_unrelated$run_id]

head(read.table(files[1], header=TRUE, check.names=FALSE)) ##checking the first few lines
tmp <- bplapply(files, read.table, header=TRUE, check.names=FALSE)
table(unlist(lapply(tmp, nrow))) ##checking all the same length
for(i in 2:length(tmp)) { ##checking all genes same position
  if(!all.equal(tmp[[1]]$gene, tmp[[i]]$gene))
    message("not equal:", i)
}
rownames <- as.character(tmp[[1]]$feature)
colnames <- gsub("\\.tsv$", "", basename(files))
countsMat <- matrix(nrow=nrow(tmp[[1]]), ncol=length(tmp),
                 dimnames=list(rownames, colnames))
for(i in 1:length(tmp)) countsMat[,i] <- tmp[[i]][,2]
countsMat[1:5, 1:5]
mid <- match(colnames(countsMat), freeze2_unrelated$run_id)
colnames(countsMat) <- freeze2_unrelated$uuid[mid]
```

## Contrust SummarizedExperiment for the entire set ##
```{r, BIOS_SE, eval=FALSE}
# remove unassigned counts
countsMat <- countsMat[grep("^Unassigned_", rownames(countsMat), invert=T),]

# prepare annotation
library(TxDb.Hsapiens.Ensembl.v91)
rowRanges <- genes(TxDb.Hsapiens.Ensembl.v91)

rownames(freeze2_unrelated) <- freeze2_unrelated$uuid

rowRanges
head(freeze2_unrelated)
countsMat[1:5, 1:5]

# make SE
counts <- makeSE(countsMat, freeze2_unrelated[,-1], rowRanges, 
                 author="D. Cats", dbVersion=mdbVersion(), note=paste0(
  "A maximal set of unrelated individuals for which both good quality RNA",
  "sequencing and DNA methylation data could be generated."))

save(counts, file = file.path(VM_BASE_DATA, "RNASeq/v2.1.3/gene_read/", 
                              "rnaSeqData_ReadCounts_BIOS_Freeze2_unrelated_GRCh38.RData"),
     compress = "xz")

```

## Contrust SummarizedExperiment per biobank ##
```{r, biobank_SE, eval=FALSE}
tmp <- counts
table(counts$biobank_id)
for(biobank in unique(tmp$biobank_id)) {
  counts <- tmp[, tmp$biobank_id == biobank]
  print(table(counts$biobank_id))
  save(counts, file = file.path(VM_BASE_DATA, "RNASeq/v2.1.3/gene_read/", paste0(
    "rnaSeqData_ReadCounts_", biobank, "_Freeze2_unrelated_GRCh38.RData")),
    compress = "xz")
}
```
