---
title: "Create DNAm data sets"
output:
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
bibliography: BBMRIomics.bib
---

```{r, child="_setup_devel.Rmd"}
```

# Prepare sample metadata #

Preprocess metadata including phenotype harmonization.

```{r}
library(BBMRIomics)
colData <- read.table(file=file.path(VM_BASE_ANALYSIS, "BBMRIomics/BBMRIomics/data", "freeze2_dnam_identifiers.csv"), header=TRUE, sep=",", as.is=TRUE)

samplesheets <- getView("methylationSamplesheet", usrpwd=RP3_MDB_USRPWD, url=RP3_MDB)
samplesheets <- samplesheets[!duplicated(samplesheets$run_id),]
colData <- merge(colData, samplesheets, by=c("run_id", "ids"))

##harmonize phenotypes
source(file.path(path.package("BBMRIomics"), "scripts/Phenotype_Helpers.R"), verbose=FALSE)
phenotypes <- cleanPhenotypes()

##add phenotypes
colData <- merge(colData, phenotypes, by="ids", all.x=TRUE)

##add genotype ids
imputation <- getView("getImputations", usrpwd=RP3_MDB_USRPWD, url=RP3_MDB)
hrc <- subset(imputation, imputation_reference == "HRC")
colData <- merge(colData, hrc[, c("ids", "imputation_id")], by="ids", all.x=TRUE)

##add uuid and gonl_id
ids <- getView("getIds", usrpwd=RP3_MDB_USRPWD, url=RP3_MDB)
colData <- merge(colData, ids[, c("ids", "gonl_id")], by="ids", all.x=TRUE)

##ordering
first <- match(c("ids", "uuid", "biobank_id", "gonl_id", "run_id"), colnames(colData))

colData <- cbind(colData[, first], colData[, -first])
rownames(colData) <- colData$uuid

dim(colData)
head(colData)

table(!is.na(colData$gonl_id))
table(!is.na(colData$imputation_id))
table(!(is.na(colData$imputation_id) & is.na(colData$gonl_id)))

```

# Creating the `SummarizedExperiments` #

Here we follow the approach described at
[DNAmArray](https://molepi.github.io/DNAmArray_workflow/) for
preprocessing and quality control.

```{r}
library(BiocParallel)
library(DNAmArray)
library(impute)
register(MulticoreParam(10, log=TRUE))
path450k <- file.path(VM_BASE_DATA, "IlluminaHumanMethylation450k")

##for(biobank in unique(colData$biobank_id)) {
biobank <- "LL"
targets <- subset(colData, biobank_id == biobank)
targets$Basename <- with(targets,
                         file.path(path450k, "raw", Sentrix_Barcode,
                                   paste(Sentrix_Barcode, Sentrix_Position, sep="_")))
##reading the data
RGset <- read.metharray.exp.par(targets, verbose=TRUE, extended=TRUE)
##normalize
pc <- screeplot(RGset)
GRset <- preprocessFunnorm.DNAmArray(RGset, nPCs=3, keepCN=FALSE)
rm(pc)
gc() #remove and clean
##filter
RGset <- probeFiltering(RGset)
##store M-values
mvalues <- reduce(GRset, RGset, what="M")
sink("/dev/null") ##to ignore impute.knn output
mvalues <- impute.knn(as.matrix(mvalues))$data
sink()

mid <- match(colnames(mvalues), colData$run_id)
colnames(mvalues) <- colData$uuid[mid]
rowData <- getPlatform(platform="HM450", genome="hg19")
mvalues <- makeSE(mvalues, colData[,-1], rowData, note="No chen probe filtering, Functional normalization, knn imputed")
save(mvalues, file=file.path(path450k, "450k", biobank, paste0("methData_Mvalues_", biobank, "_F2_cleaned.RData")))
rm(mvalues)
gc() #remove and clean
##store beta-values
betas <- reduce(GRset, RGset, what="beta")
sink("/dev/null") ##to ignore impute.knn output
betas <- impute.knn(as.matrix(betas))$data
sink()
mid <- match(colnames(betas), colData$run_id)
colnames(betas) <- colData$uuid[mid]
betas <- makeSE(betas, colData[,-1], rowData, note="No chen probe filtering, Functional normalization, knn imputed")
save(betas, file=file.path(path450k, "450k", biobank, paste0("methData_Betas_", biobank, "_F2_cleaned.RData")))
rm(betas)
rm(GRset)
rm(RGset)
gc() ##remove and clean
##}

```

# Create one `SummarizedExperiment` containg all cohorts #

Here we merge the `SummarizedExperiment` of the separate cohorts in one
single `SummarizedExperiment`.

> NOTE: direct use of `cbind` currently doesn't work with `GRanges`
> containing `DNAStringSets` problem has been reported to
> Bioconductor-developers who proposed to fix it!

```{r, eval=FALSE}
path450k <- file.path(VM_BASE_DATA, "IlluminaHumanMethylation450k")
biobank <- "PAN"
load(file.path(path450k, "450k", biobank, paste0("methData_Mvalues_", biobank, "_F2_cleaned.RData")))
b <- mvalues
library(GenomicRanges)
for(biobank in c("CODAM", "LLS", "LL", "RS", "NTR")) {
    load(file.path(path450k, "450k", biobank, paste0("methData_Mvalues_", biobank, "_F2_cleaned.RData"))) 
    mvalues <- subsetByOverlaps(mvalues, b)
    b <- subsetByOverlaps(b, mvalues)
    rowData(mvalues) <- NULL
    b <- SummarizedExperiment::cbind(b, mvalues)
    gc()
    print(b)
}
mvalues <- b
save(mvalues, file=file.path(path450k, "450k", paste0("methData_Mvalues_BIOS_F2_cleaned.RData")))
##can we reduce the size?
tools::checkRdaFiles(file.path(path450k, "450k", paste0("methData_Mvalues_BIOS_F2_cleaned.RData")))
```
