---
title: "Available Data"
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
---

```{r, child="_setup.Rmd"}
```

## Generated Omic data ##

***brief description***

## Available datasets ##

***brief description***

## Loading the data ##

Load a specific data set with `data(name_dataset)`. The **BBMRIomics**
global variable `DATASETS` lists all available data sets. Once loaded
check the name of the loaded data with `ls()` and view its content by
just typing the name of the dataset in the console this will
automatically call the buildin `show`-method and summarizes the
content of the loaded dataset.

```{r, datasets}
library(BBMRIomics)
data(package="BBMRIomics")
```

```{r, loading}
data(methData_Betas_CODAM_F2_cleaned)
ls()
betas
```

The datasets are **R/Bioconductor**
**SummarizedExperiment**-objects. The **SummarizedExperiment** class
is used to store rectangular matrices of experimental results; a data
container storing feature data like gene counts or betas-values,
annotation on the features, gene identifiers or genomic location, as
well as annotation on the samples, i.e. phenotypes and metadata.

The advantage of having all information in one object should be
clear. Subsetting the object will automatically subset the annotation
or sample information. 

See for more information on **SummarizedExperiments**:
1. [Vignette](http://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html)
2. [Tutorial](http://bioconductor.org/help/course-materials/2012/BiocEurope2012/SummarizedExperiment.pdf)
3. [Paper](http://www.nature.com/nmeth/journal/v12/n2/full/nmeth.3252.html)






