---
title: "Available Data"
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
---

```{r, child="_setup.Rmd"}
```

## Generated Omic data ##

***brief description***

## Available datasets ##

***brief description***

## Loading the data ##

Load a specific data set with `data(name_dataset)`. The **BBMRIomics**
global variable `DATASETS` lists all available data sets. Once loaded
check the name of the loaded data with `ls()` and view its content by
just typing the name of the dataset in the console this will
automatically call the buildin `show`-method and summarizes the
content of the loaded dataset.

```{r, datasets}
library(BBMRIomics)
data(package="BBMRIomics")
```

```{r, loading}
data(methData_Betas_CODAM_F2_cleaned)
ls()
betas
```

## SummarizedExperiment ##

The datasets are stored in the **R/Bioconductor**
**SummarizedExperiment**-objects.

*Brief summary from the* **SummarizedExperiment** *vignette:*

The `SummarizedExperiment` class is used to store rectangular matrices
of experimental results, which are commonly produced by sequencing and
microarray experiments. Each object stores observations of one or more
samples, along with additional meta-data describing both the
observations (features) and samples (phenotypes).

A key aspect of the `SummarizedExperiment` class is the coordination
of the meta-data and assays when subsetting. For example, if you want
to exclude a given sample you can do for both the meta-data and assay
in one operation, which ensures the meta-data and observed data will
remain in sync. Improperly accounting for meta and observational data
has resulted in a number of incorrect results and retractions so this
is a very desirable property.

`SummarizedExperiment` is in many ways similar to the historical
`ExpressionSet`, the main distinction being that `SummarizedExperiment` is
more flexible in it’s row information, allowing both `GRanges` based as
well as those described by arbitrary `DataFrame`s. This makes it ideally
suited to a variety of experiments, particularly sequencing based
experiments such as RNA-Seq and ChIp-Seq.

See for more information on the **SummarizedExperiment** datastructure:

1. [Vignette](http://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html)
2. [Tutorial](http://bioconductor.org/help/course-materials/2012/BiocEurope2012/SummarizedExperiment.pdf)
3. [Paper](http://www.nature.com/nmeth/journal/v12/n2/full/nmeth.3252.html)





