---
title: "450k quality control"
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
bibliography: BBMRIomics.bib
---

```{r, child="_setup_devel.Rmd"}
```

## Sample-level quality control ##

The package
[**MethylAid**](http://bioconductor.org/packages/MethylAid/) is used
for sample-level quality
control. [**MethylAid**](http://bioconductor.org/packages/MethylAid/)
is specially designed for quality control of large sets of Illumina
Human DNA methylation array data sets e.g., epigenomewide association
studies (EWAS) using the 450k or 850k (EPIC) arrays. Extracting
intensities from IDAT files can be done in batches and/or in parallel
to reduce memory load and/or overcome long run-times. It requires two
function calls in going from IDAT files to launch the interactive web
application; `summarize` and `visualize`. For more information see
[@Iterson2014].

### Creating targets-files ###

```{r, eval=FALSE}
path450k <- file.path(VM_BASE_DATA, "IlluminaHumanMethylation450k")
samplesheets <- getSQLview("methylationSamplesheet")
samplesheets <- samplesheets[!duplicated(samplesheets$run_id),]
samplesheets <- samplesheets[order(samplesheets$ids),]
head(samplesheets)
dim(samplesheets)
for(biobank in c("CODAM", "LL", "LLS", "NTR", "PAN", "RS"))
    write.table(subset(samplesheets, biobank_id == biobank),
                file = file.path(path450k, "450k", biobank, "sample_sheet.txt"),
                sep=",", row.names=FALSE, quote=FALSE)
samplesheets$Basename <- with(samplesheets,
                              file.path(path450k, "raw", Sentrix_Barcode,
                                        paste(Sentrix_Barcode, Sentrix_Position, sep="_")))
```

Now a check is performed to see if all idat-files exists or if we
missed one on the VM.


```{r, eval=FALSE}
idats <- c(paste0(samplesheets$Basename, "_Red.idat"),
           paste0(samplesheets$Basename, "_Grn.idat"))
exists <- file.exists(idats)
table(exists)
files <- dir(file.path(path450k, "raw"), recursive=TRUE, full.names=TRUE)
setdiff(files, idats)
setdiff(idats, files)
```

### Summarize ###

Now we are ready to run **MethylAid** summarization.

For each biobank we extract all quality control probe intensities
using the **MethylAid** function `summarize`. Since, we
have a VM with multiple cores we can do the summarization in
parallel. For each biobank we generate an `RData`-file called
**MethylAid** and store it in the biobanks data directory.

```{r, eval=FALSE}
register(MulticoreParam(10, log=TRUE))
tmp <- sapply(c("CODAM", "LL", "LLS", "NTR", "PAN", "RS"), function(biobank) {
    samplesheet <- samplesheets[samplesheets$biobank_id == biobank,]
    summarize(samplesheet, batchSize = 100, BPPARAM = bpparam(),
              file=file.path(path450k, "450k", biobank, "MethylAid"))
})
```

See Apps for the interactive sample-level quality control.

