---
title: "Phenotyping using RNAseq data"
output:
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
bibliography: BBMRIomics.bib
---

```{r, child="_setup.Rmd"}
```

# Intro #

This document describes how we try to impute some missing phenotypes,
specifically, sex, age, smoking status and white blood cell
composition.

These phenotypes have missing values for ten to a few hundred
individuals, a few percentages of the whole data, leaving enough
individuals to build accurate predictors for these phenotypes.

The [**omicsPrint**](https://github.com/molepi/omicsPrin) packages is
developed to ease the construction of predictive models using
high-dimensional data. It provides convenient interfaces to the
[**glmnet**](https://cran.r-project.org/web/packages/glmnet/index.html),
[**pls**](https://cran.r-project.org/web/packages/pls/index.html) and
[**gbm**] packages (potentially more can be added in the future),
i.e. in a unified framework.


Different response types can be handle by all algorithms; categorical
(Sex/Smoking Status) and continuous (Age, white blood cell
composition).

The currently available algorithms are:

 1. ridge, lasso and elastic net (**glmnet**)
 2. partial-least squares (**pls**)
 3. gradient boosting (**gbm**)


$\alpha = 0$ ridge, $\alpha = 1$ lasso, $\alpha = 0.5$ elastic-net 

ridge- or lasso-penalty is optimized using interal cross-validation

continuous is linear regression (family=gaussian)

categorical repsonse use logistic regression (with family binomial or multinomial)


detect wrongly labeled samples



Here we define a few helper functions:

```{r}
library(omicsPrint)

preprocess <- function(sumExp) {
    require(SummarizedExperiment)
    require(edgeR)

    y <- DGEList(assays(sumExp)$data)

    message("[preprocess] drop features not expressed")
    ##https://www.bioconductor.org/packages/devel/bioc/vignettes/limma/inst/doc/usersguide.pdf
    A <- rowSums(y$counts)
    isexpr <- A > 50
    y <- y[isexpr,]

    message("[preprocess] normalize")
    y <- calcNormFactors(y) ##normalize takes some time

    message("[preprocess] calculate log counts per million")
    y <- cpm(y, log=TRUE)

    sumExp <- sumExp[isexpr,]
    assays(sumExp)$data <- y ##replace data with filtered subset
    sumExp
}

validate <- function(phenotype, features, train.frac = 2/3, methods = c("ridge", "elastic-net", "partial-least-squares", "lasso", "gradient-boosting"), ntop = c(50, 100, 250, 500, 1000, 2500, 5000), nrep = 25, verbose=TRUE) {
    require(omicsPrint)
    require(reshape2)
    require(BiocParallel)
    validations <- c()
    for(ntopi in ntop) {
        reps <- bplapply(1:nrep, function(x) phenotyping(phenotype, features, train.frac=train.frac, methods = methods, ntop=ntopi, verbose=verbose)$validation)
        reps <- do.call('rbind', reps)
        validation <- melt(reps, varnames=c("method", "measure"))
        validation$ntop <- ntopi
        validations <- rbind(validations, validation)
        gc()
    }
    validations
}


```

# Inspect missing data #

```{r}
library(BBMRIomics)
data(rnaSeqData_ReadCounts_BIOS_cleaned)
covariates <- as.data.frame(colData(counts)[, c("Sex", "Sampling_Age", "Smoking", "Lymph_Perc","Mono_Perc", "Eos_Perc", "Neut_Perc")])
##missing covariate summary
missingCov <- apply(covariates, 2, function(x) round(100*table(is.na(x))/length(x),2))
missingCov
##missing sample summary
missingSam <- apply(covariates, 1, function(x) round(100*sum(is.na(x))/length(x),2))
table(missingSam)
```

# Predicting Sex #

## Model selection and Validation ##

First we need to find out which algorithm and which number of features
to use. 


```{r}
data(rnaSeqData_ReadCounts_BIOS_cleaned)
metadata(counts)$formula <- ~0+Sex

dim(counts)
cnts <- preprocess(counts)
dim(cnts)

phenotype <- unlist(get_all_vars(metadata(cnts)$formula, data=colData(cnts)))
features <- assays(cnts)$data
table(phenotype, useNA="always")

library(BiocParallel)
register(MulticoreParam(5))
validation <- validate(phenotype, features, methods = c("ridge", "elastic-net", "lasso"), ntop = c(5, 10, 25, 50, 100), nrep = 25, verbose=FALSE)

library(ggplot2)
library(ggsci)
gp <- ggplot(subset(validation, measure %in% c("accuracy (Overall)", "f1 (Male)", "f1 (Female)")),
             aes(x=as.factor(ntop), y=value, col=method))
gp <- gp + geom_boxplot()
gp <- gp + facet_wrap(~measure, scales="free", nrow=3, ncol=1)
gp <- gp + xlab("Number of features") + ylab("") + ggtitle("Validation of predictors for `Sex`")
gp + theme_bw() + scale_color_npg() + theme(legend.position = "bottom", legend.direction = "horizontal")

```


## Predict missing ##

```{r}
phenotyped <- phenotyping(phenotype, features, methods = "lasso", ntop=50, verbose=TRUE)
phenotyped$validation
predicted <- phenotyped$predicted
names(predicted) <- colnames(features)

write.table(predicted, file="/virdir/Scratch/RP3_analysis/imputed_phenotypes/Sex_RNAseq.csv",
            sep=",", quote=FALSE)

```

> TODO add diagnostic plots!

## Inspect selected features ##

It could be interesting to see which genes are the top genes selected
for predicting Age.

> this does not include e.g., the lasso selection

```{r}
library(org.Hs.eg.db)
map <- select(org.Hs.eg.db, keys = names(phenotyped$top), keytype="ENSEMBL", columns="SYMBOL")
top <- merge(map, phenotyped$top, by.x="ENSEMBL", by.y="row.names")
top <- top[order(abs(top$y), decreasing=TRUE),]
top

mid <- match(top$ENSEMBL, rownames(features))
y <- features[mid,]
rownames(y) <- top$SYMBOL
rownames(y)[is.na(top$SYMBOL)] <- top$ENSEMBL[is.na(top$SYMBOL)]
colnames(y) <- NULL

library(ComplexHeatmap)
sex <- colData(cnts)$Sex
sex[is.na(sex)] <- "Unknown"
ha <- HeatmapAnnotation(df = data.frame(Sex = factor(sex)), col = list(Sex = c("Male" =  "blue", "Female" = "pink", "Unknown" = "grey")))
Heatmap(y, top_annotation = ha, name="log CPM", row_names_gp = gpar(fontsize = 8))
```


## Detect outliers ##

Outlier or missclassified sample could potentially be wrongly labeled.

Since, we use a subset for training it could be accidentally be
missclassified. Therefore, we run multiple models and collected the
number and frequencies of missclassified samples.


```{r}
library(BiocParallel)
register(MulticoreParam(10))
nrep <- 25
predicted <- bplapply(1:nrep, function(x) phenotyping(phenotype, features, methods = "ridge", ntop=50, verbose=FALSE)$predicted)
predicted <- do.call('cbind', predicted)

missclassified <- (predicted != phenotype)[!is.na(phenotype), ]
rownames(missclassified) <- colnames(features)[!is.na(phenotype)]
missclassified <- missclassified[rowSums(missclassified) > 0,]
100*rowSums(missclassified)/nrep

```

# Predicting Age #

## Model selection and Validation ##

First we need to find out which algorithm and which number of features
to use:

```{r, eval=FALSE}

metadata(cnts)$formula <- ~0+Sampling_Age

phenotype <- unlist(get_all_vars(metadata(cnts)$formula, data=colData(cnts)))
features <- assays(cnts)$data
summary(phenotype)

library(BiocParallel)
register(MulticoreParam(5))
validation <- validate(phenotype, features, methods = c("ridge", "elastic-net", "lasso", "gradient-boosting", "partial-least-squares"),
                       ntop = c(100, 250, 500, 1000, 2500, 5000), nrep = 25, verbose=FALSE)

library(ggplot2)
library(ggsci)
gp <- ggplot(validation, aes(x=as.factor(ntop), y=value, col=method))
gp <- gp + geom_boxplot()
gp <- gp + facet_wrap(~measure, scales="free", nrow=3, ncol=1)
gp <- gp + xlab("Number of features") + ylab("") + ggtitle("Validation of predictors for `Age`")
gp + theme_bw() + scale_color_npg() + theme(legend.position = "bottom", legend.direction = "horizontal")

```

some diagnostic plots:

```{r, eval=FALSE}
phenotyped <- phenotyping(phenotype, features, methods = "lasso", ntop=5000, verbose=TRUE)
predicted <- phenotyped$predicted
names(predicted) <- colnames(features)

plot(predicted, phenotype, xlab="predicted (Age)", ylab="reported (Age)",
     xlim=c(0, 100), ylim=c(0, 100),
     pch=16, col = adjustcolor(1, alpha.f = 0.3))
abline(0, 1, col=2)
grid(col=1)

residuals <- predicted-phenotype
plot(predicted[!is.na(phenotype)], residuals[!is.na(phenotype)],
     xlab="fitted", ylab="residual", pch=16, col = adjustcolor(1, alpha.f = 0.3))
grid(col=1)

qqnorm(residuals)
qqline(residuals)

```

## Predict missing ##

```{r, eval=FALSE}
phenotyped <- phenotyping(phenotype, features, methods = "lasso", ntop=5000, verbose=TRUE)
predicted <- phenotyped$predicted
names(predicted) <- colnames(features)
predicted[is.na(phenotype)]

```

## Inspect selected features ##


```{r, eval=FALSE}
library(org.Hs.eg.db)
map <- select(org.Hs.eg.db, keys = names(phenotyped$top), keytype="ENSEMBL", columns="SYMBOL")
top <- merge(map, phenotyped$top, by.x="ENSEMBL", by.y="row.names")
head(top[order(abs(top$y), decreasing=TRUE),])

library(irlba)
y <- features[rownames(features) %in% names(phenotyped$top),]
pc <- prcomp_irlba(t(y))
summary(pc)

cols <- rainbow(10)[cut(phenotype, breaks=10)]
plot(pc$x, col=adjustcolor(cols, alpha.f=0.3), pch=16)

```


## Detect outliers ##

```{r, eval=FALSE}

library(BiocParallel)
register(MulticoreParam(10))
nrep <- 100
predicted <- bplapply(1:nrep, function(x) phenotyping(phenotype, features, methods = "lasso", ntop=5000, verbose=FALSE)$predicted)
predicted <- do.call('cbind', predicted)

residuals <- predicted - phenotype

popo <- which(rowSums(abs(residuals) > mean(residuals, na.rm=TRUE) + 3*sd(residuals, na.rm=TRUE)) > 10)

library(matrixStats)
mns <- rowMeans(predicted)
sds <- rowSds(predicted)
col <- rep(1, length(mns))
col[popo] <- 2
plot(phenotype, mns, pch=16, cex=sds, col=adjustcolor(col, alpha.f=0.3), bty="n")
abline(0, 1, col=4, lwd=2, lty=2)
grid(col=1)


```


# Predict white blood cell composition #


## Model fitting and model selection ##

```{r, wbccfit}
library(BBMRIomics)
data(rnaSeqData_ReadCounts_BIOS_cleaned)

metadata(counts)$formula <- ~0 + Lymph_Perc + Mono_Perc + Eos_Perc + Neut_Perc

dim(counts)
cnts <- preprocess(counts)
dim(cnts)

phenotypes <- get_all_vars(metadata(cnts)$formula, data=colData(cnts))
features <- assays(cnts)$data

complete <- apply(phenotypes, 1, function(x) all(!is.na(x)))
table(complete)

head(phenotypes[complete,])

trainId <- sample(which(complete), size = (2/3)*sum(complete))
testId <- setdiff(which(complete), trainId)

library(pls)
pls.options(parallel = 5)
predictor <- plsr(phenotypes~data, ncomp = 50, data = list(phenotypes = as.matrix(phenotypes[trainId,]), data=t(features[, trainId])), validation = "CV", keep.model = TRUE)

str(predictor)
summary(predictor)
cumsum(explvar(predictor))
barplot(cumsum(explvar(predictor)), las = 2, ylab = "Cumulative Variance Explained")

validationplot(predictor, val.type = "R2", ncomp = 1:50)
validationplot(predictor, val.type = "RMSEP", ncomp = 1:50)

```


## Evaluate predictions ##

```{r, wbcceval}

predicted <- predict(predictor, newdata = list(data=t(features[, testId])) , ncomp = 20)

library(reshape2)
data <- cbind(melt(phenotypes[testId,], value.name="measured"),
              predicted=melt(predicted[,,1])[,3])

##correlation plot
library(ggplot2)
library(ggsci)
gp <- ggplot(data, aes(x=measured, y=predicted))
gp <- gp + geom_point() + geom_abline(intercept=0, slope=1, col=2)
gp <- gp + facet_wrap(~variable, scales="free")
gp + theme_bw() + scale_color_npg(alpha=0.3)

##bland altman plot
tdata <- data
tdata[,2] <- rowMeans(data[,-1])
tdata[,3] <- data[,2] - data[,3]
colnames(tdata)[2:3] <- c("Average", "Difference")

gp <- ggplot(tdata, aes(x=Average, y=Difference))
gp <- gp + geom_point(alpha=0.1) + geom_hline(yintercept=0, col=2)
gp <- gp + facet_wrap(~variable, scales="free")
gp + theme_bw() + scale_color_npg()

```



```{r, wbccpred}

predicted <- predict(predictor, newdata = list(data=t(features)), ncomp = 20)[,,1]
hist(rowSums(predicted))
dim(predicted)
rownames(predicted) <- colnames(features)
head(predicted)

write.table(predicted, file="/virdir/Scratch/RP3_analysis/imputed_phenotypes/WBCC_RNAseq.csv",
            sep=",", quote=FALSE)


```
