---
title: "Phenotyping using 450K DNA methylation data"
output:
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
bibliography: BBMRIomics.bib
---

```{r, child="_setup.Rmd"}
```

Here we define a few helper functions:

```{r}
library(omicsPrint)

preprocess <- function(sumExp, dropXY=FALSE) {
    require(SummarizedExperiment)
    require(DNAmArray)  

    message("[preprocess] drop Zhou probes...")
    data(hm450.manifest.pop.GoNL) ##From DNAmArray
    
    hm450.manifest.pop.GoNL
    hm450.manifest.pop.GoNL <- hm450.manifest.pop.GoNL[!is.na(hm450.manifest.pop.GoNL$MASK.general.GoNL) &
                                                   hm450.manifest.pop.GoNL$MASK.general.GoNL == TRUE, ]

    sumExp <- sumExp[!(names(sumExp) %in% names(hm450.manifest.pop.GoNL)),]

    if(dropXY) {
        message("[preprocess] drop XY probes...")
        sumExp <- sumExp[!(seqnames(sumExp) %in% c("chrX", "chrY")),]
    }
    
    sumExp
}

validate <- function(phenotype, features, train.frac = 2/3, methods = c("ridge", "elastic-net", "pls", "lasso", "gbm"), ntop = c(50, 100, 250, 500, 1000, 2500, 5000), nrep = 25, verbose=TRUE) {
    require(omicsPrint)
    require(reshape2)
    require(BiocParallel)
    validations <- c()
    for(ntopi in ntop) {
        if(verbose)
            message("Running with ", ntopi, " features...")
        reps <- bplapply(1:nrep, function(x) phenotyping(phenotype, features, train.frac=train.frac, methods = methods, ntop=ntopi, verbose=verbose)$validation)
        reps <- do.call('rbind', reps)
        validation <- melt(reps, varnames=c("method", "measure"))
        validation$ntop <- ntopi
        validations <- rbind(validations, validation)
        gc()
    }
    validations
}

path <- "/virdir/Scratch/RP3_analysis/imputed_phenotypes"

```

# Inspect missing data #

```{r}
library(BBMRIomics)
data(methData_Betas_BIOS_F2_cleaned)
covariates <- as.data.frame(colData(betas)[, c("Sex", "Sampling_Age", "Smoking", "Lymph_Perc","Mono_Perc", "Eos_Perc", "Neut_Perc")])

##missing covariate summary
missingCov <- apply(covariates, 2, function(x) round(100*table(is.na(x))/length(x),2))
missingCov
apply(covariates, 2, function(x) table(is.na(x)))
##missing sample summary
missingSam <- apply(covariates, 1, function(x) round(100*sum(is.na(x))/length(x),2))
table(missingSam)

```

# Predicting Sex #

## Model selection and Validation ##

First we need to find out which algorithm and the number of features
lead to an optimal prediction.


```{r}

dim(betas)
bts <- preprocess(betas)
dim(bts)
metadata(bts)$formula <- ~0+Sex

phenotype <- unlist(get_all_vars(metadata(bts)$formula, data=colData(bts)))
features <- assays(bts)$data
table(phenotype, useNA="always")

library(BiocParallel)
register(MulticoreParam(10))
validation <- validate(phenotype, features, methods = c("ridge", "elastic-net", "lasso"), ntop = c(5, 10, 25, 50, 100), nrep = 25, verbose=FALSE)

library(ggplot2)
library(ggsci)
gp <- ggplot(subset(validation, measure %in% c("accuracy (Overall)", "f1 (Male)", "f1 (Female)")),
             aes(x=as.factor(ntop), y=value, col=method))
gp <- gp + geom_boxplot()
gp <- gp + facet_wrap(~measure, scales="free", nrow=3, ncol=1)
gp <- gp + xlab("Number of features") + ylab("") + ggtitle("Validation of predictors for `Sex`")
gp + theme_bw() + scale_color_npg() + theme(legend.position = "bottom", legend.direction = "horizontal")

ggsave(file.path(path, "Sex_DNAm_eval.eps"))

```


## Predict missing ##

```{r}
phenotyped <- phenotyping(phenotype, features, methods = "elastic-net", ntop=10, verbose=TRUE)
phenotyped$validation
predicted <- phenotyped$predicted
rownames(predicted) <- colnames(features)
write.table(predicted, file=file.path(path, "Sex_DNAm.csv"), sep=",", quote=FALSE)

```

> TODO add diagnostic plots!

## Inspect selected features ##

It could be interesting to see which genes are the top genes selected
for predicting Age.

> this does not include e.g., the lasso selection

```{r, eval=FALSE}
top <- bts[rownames(bts) %in% names(phenotyped$top),]
rowRanges(top)
feats <- assays(top)$data
colnames(feats) <- NULL
library(ComplexHeatmap)
sex <- colData(top)$Sex
sex[is.na(sex)] <- "Unknown"
setEPS()
postscript(file.path(path, "Sex_DNAm_feats.eps"))
Heatmap(feats, top_annotation = ha, name="beta-value", row_names_gp = gpar(fontsize = 8))
dev.off()

```


## Detect outliers ##

Outlier or missclassified sample could potentially be wrongly labeled.

Since, we use a subset for training it could be accidentally be
missclassified. Therefore, we run multiple models and collected the
number and frequencies of missclassified samples.


```{r, eval=FALSE}
library(BiocParallel)
register(MulticoreParam(15))
nrep <- 25
predicted <- bplapply(1:nrep, function(x) phenotyping(phenotype, features,  methods = "elastic-net", ntop=10, verbose=FALSE)$predicted)
predicted <- do.call('cbind', predicted)
missclassified <- (predicted != phenotype)[!is.na(phenotype), ]
rownames(missclassified) <- colnames(features)[!is.na(phenotype)]
missclassified <- missclassified[rowSums(missclassified) > 0,]
100*rowSums(missclassified)/nrep

```

# Predicting Age #

## Model selection and Validation ##

First we need to find out which algorithm and which number of features
to use:

```{r}
metadata(bts)$formula <- ~0+Sampling_Age
phenotype <- unlist(get_all_vars(metadata(bts)$formula, data=colData(bts)))
features <- assays(bts)$data
summary(phenotype)

library(BiocParallel)
register(MulticoreParam(10))
validation <- validate(phenotype, features, methods = c("ridge", "elastic-net", "lasso", "pls"),
                       ntop = c(100, 250, 500, 1000, 2500, 5000), nrep = 25, verbose=TRUE)

library(ggplot2)
library(ggsci)
gp <- ggplot(validation, aes(x=as.factor(ntop), y=value, col=method))
gp <- gp + geom_boxplot()
gp <- gp + facet_wrap(~measure, scales="free", nrow=3, ncol=1)
gp <- gp + xlab("Number of features") + ylab("") + ggtitle("Validation of predictors for `Age`")
gp + theme_bw() + scale_color_npg() + theme(legend.position = "bottom", legend.direction = "horizontal")

ggsave(file.path(path, "Age_DNAm_eval.eps"))

```

some diagnostic plots:

```{r}
phenotyped <- phenotyping(phenotype, features, methods = "pls", ntop=5000, verbose=TRUE)
predicted <- phenotyped$predicted
rownames(predicted) <- colnames(features)

type <- rep("train", length(phenotype))
type[is.na(phenotype)] <- "unknown"
type[phenotyped$testid] <- "test"

phenotype[is.na(phenotype)] <- predicted[is.na(phenotype)]

d <- data.frame(predicted=predicted[,1], measured=phenotype, type=factor(type))

gp <- ggplot(d, aes(x=predicted, y=measured, col=type))
gp <- gp + geom_point() + geom_abline(intercept=0, slope=1, col=1) + ggtitle("Age prediction")
gp + theme_bw() + scale_color_npg() + theme(legend.position = "bottom", legend.direction = "horizontal")
ggsave(file.path(path, "Age_DNAm_diagn.eps"))

```

## Predict missing ##

```{r}
phenotype <- unlist(get_all_vars(metadata(bts)$formula, data=colData(bts)))
phenotyped <- phenotyping(phenotype, features, methods = "pls", ntop=5000, verbose=TRUE)
predicted <- phenotyped$predicted
rownames(predicted) <- colnames(features)
write.table(predicted, file=file.path(path, "Age_DNAm.csv"), sep=",", quote=FALSE)
```

## Inspect selected features ##


```{r, eval=FALSE}
library(irlba)
y <- features[rownames(features) %in% names(phenotyped$top),]
pc <- prcomp_irlba(t(y))
summary(pc)

Age <- cut(phenotype, breaks=10)
cols <- rainbow(10)[Age]
plot(pc$x, col=adjustcolor(cols, alpha.f=0.3), pch=16)

library(ComplexHeatmap)
sex <- colData(top)$Sex
sex[is.na(sex)] <- "Unknown"

ha <- HeatmapAnnotation(df = data.frame(Age = factor(age)), col = list(Sex = c("Male" =  "blue", "Female" = "pink", "Unknown" = "grey")))

setEPS()
postscript(file.path(path, "Sex_DNAm_feats.eps"))

Heatmap(y, name="beta-value", row_names_gp = gpar(fontsize = 8))

dev.off()


```


## Detect outliers ##

```{r, eval=FALSE}

library(BiocParallel)
register(MulticoreParam(10))
nrep <- 100
predicted <- bplapply(1:nrep, function(x) phenotyping(phenotype, features, methods = "lasso", ntop=5000, verbose=FALSE)$predicted)
predicted <- do.call('cbind', predicted)

residuals <- predicted - phenotype

popo <- which(rowSums(abs(residuals) > mean(residuals, na.rm=TRUE) + 3*sd(residuals, na.rm=TRUE)) > 10)

library(matrixStats)
mns <- rowMeans(predicted)
sds <- rowSds(predicted)
col <- rep(1, length(mns))
col[popo] <- 2
plot(phenotype, mns, pch=16, cex=sds, col=adjustcolor(col, alpha.f=0.3), bty="n")
abline(0, 1, col=4, lwd=2, lty=2)
grid(col=1)


```


# Predict white blood cell composition #


## Model fitting and model selection ##

```{r, wbccfit}

metadata(bts)$formula <- ~0 + Lymph_Perc + Mono_Perc + Eos_Perc + Neut_Perc

phenotypes <- get_all_vars(metadata(bts)$formula, data=colData(bts))
features <- assays(bts)$data

complete <- apply(phenotypes, 1, function(x) all(!is.na(x)))
table(complete)

head(phenotypes[complete,])

trainId <- sample(which(complete), size = (2/3)*sum(complete))
testId <- setdiff(which(complete), trainId)

length(trainId)
length(testId)

Sentrix_Position <- colData(bts)$Sentrix_Position
batches <- model.matrix(~Sentrix_Position)

library(pls)
pls.options(parallel = 5)
predictor <- plsr(log10(phenotypes+1)~batches+data, ncomp = 50,
                  data = list(phenotypes = as.matrix(phenotypes[trainId,]), batches=batches[trainId, ],
                              data=t(features[, trainId])), validation = "CV", keep.model = TRUE)
summary(predictor)
cumsum(explvar(predictor))
barplot(cumsum(explvar(predictor)), las = 2, ylab = "Cumulative Variance Explained")

validationplot(predictor, val.type = "R2", ncomp = 1:50)
validationplot(predictor, val.type = "RMSEP", ncomp = 1:50)

```


## Evaluate predictions ##

```{r, wbcceval}

predicted <- predict(predictor, newdata = list(data=t(features[, testId]), batches=batches[testId,]) , ncomp = 40)[,,1]
predicted <- 10^predicted - 1

library(reshape2)
data <- cbind(melt(phenotypes[testId,], value.name="measured"),
              predicted=melt(predicted)[,3])

##hack to get plot area square
dummy <- t(simplify2array(by(data, data$variable, function(x) range(c(x$measured, x$predicted)))))
dummy <- data.frame(variable=rep(rownames(dummy), 2), measured=as.vector(dummy), predicted=as.vector(dummy))

by(data, data$variable, function(x) round(cor(x$measured, x$predicted), 2))

##correlation plot
library(ggplot2)
library(ggsci)
gp <- ggplot(data, aes(x=measured, y=predicted))
gp <- gp + geom_point() + geom_abline(intercept=0, slope=1, col=2)
gp <- gp + geom_blank(data=dummy)
gp <- gp + facet_wrap(~variable, scales="free")
gp + theme_bw() + scale_color_npg()
ggsave(file.path(path, "WBCC_DNAm_diagn1.eps"))

##bland altman plot
tdata <- data
tdata[,2] <- rowMeans(data[,-1])
tdata[,3] <- data[,2] - data[,3]
colnames(tdata)[2:3] <- c("Average", "Difference")

gp <- ggplot(tdata, aes(x=Average, y=Difference))
gp <- gp + geom_point() + geom_hline(yintercept=0, col=2)
gp <- gp + facet_wrap(~variable, scales="free")
gp + theme_bw() + scale_color_npg()
ggsave(file.path(path, "WBCC_DNAm_diagn2.eps"))

```



```{r, wbccpred}

predicted <- predict(predictor, newdata = list(data=t(features), batches=batches) , ncomp = 40)[,,1]
predicted <- 10^predicted - 1

hist(rowSums(predicted))
dim(predicted)
rownames(predicted) <- colnames(features)
head(predicted)

write.table(predicted, file=file.path(path, "WBCC_DNAm.csv"), sep=",", quote=FALSE)


```


# Predicting Smoking #
## Model Selection and Validation ##

First we need to find out which algorithm and which number of features
to use. 

```{r}

metadata(bts)$formula <- ~0+Smoking

phenotype <- unlist(get_all_vars(metadata(bts)$formula, data=colData(bts)))
features <- assays(bts)$data
table(phenotype, useNA="always")/length(phenotype)
levels(phenotype) <- c("current smoker", "non-smoker", "non-smoker")
table(phenotype, useNA="always")/length(phenotype)

library(BiocParallel)
register(MulticoreParam(5))
validation <- validate(phenotype, features, methods = c("ridge", "elastic-net", "lasso"), ntop = c(10, 25, 50, 100, 250), nrep = 5, verbose=TRUE)

library(ggplot2)
library(ggsci)
gp <- ggplot(subset(validation, measure %in% c("accuracy (Overall)", "f1 (non-smoker)", "f1 (current smoker)")),
             aes(x=as.factor(ntop), y=value, col=method))
gp <- gp + geom_boxplot()
gp <- gp + facet_wrap(~measure, scales="free", nrow=3, ncol=1)
gp <- gp + xlab("Number of features") + ylab("") + ggtitle("Validation of predictors for `Smoking`")
gp + theme_bw() + scale_color_npg() + theme(legend.position = "bottom", legend.direction = "horizontal")

ggsave(file.path(path, "Smoking_DNAm_eval.eps"))

```

## Predict missing ##

```{r}
phenotyped <- phenotyping(phenotype, features, methods = "elastic-net", ntop=50, verbose=TRUE)
phenotyped$validation
predicted <- phenotyped$predicted
rownames(predicted) <- colnames(features)
write.table(predicted, file=file.path(path, "Smoking_DNAm.csv"), sep=",", quote=FALSE)
```

> TODO add diagnostic plots!

## Inspect selected features ##

It could be interesting to see which genes are the top genes selected
for predicting Age.

> this does not include e.g., the lasso selection

```{r, eval=FALSE}
library(org.Hs.eg.db)
map <- select(org.Hs.eg.db, keys = names(phenotyped$top), keytype="ENSEMBL", columns=c("SYMBOL", "CHR"))
top <- merge(map, phenotyped$top, by.x="ENSEMBL", by.y="row.names")
top <- top[order(abs(top$y), decreasing=TRUE),]
top

mid <- match(top$ENSEMBL, rownames(features))
y <- features[mid,]
rownames(y) <- top$SYMBOL
rownames(y)[is.na(top$SYMBOL)] <- top$ENSEMBL[is.na(top$SYMBOL)]
colnames(y) <- NULL

library(ComplexHeatmap)
smoking <- as.character(phenotype)
smoking[is.na(smoking)] <- "Unknown"
ha <- HeatmapAnnotation(df = data.frame(Smoking = factor(smoking)), col = list(Smoking = c("current smoker" =  "blue", "non-smoker" = "pink", "Unknown" = "grey")))

setEPS()
postscript(file.path(path, "Smoking_DNAm_feats.eps"))
Heatmap(y, top_annotation = ha, name="beta-value", row_names_gp = gpar(fontsize = 8))
dev.off()
```


# Concordance #


```{r, eval=FALSE}

files <- dir(path, pattern="Sex.*csv", full.names=TRUE)
d1 <- read.table(files[1], sep=",", header=TRUE)
d2 <- read.table(files[2], sep=",", header=TRUE)
d <- merge(d1, d2, by="row.names")
colnames(d)[-1] <- c("DNAm", "RNAseq")
table(d[,2], d[,3])

files <- dir(path, pattern="Age.*csv", full.names=TRUE)
d1 <- read.table(files[1], sep=",", header=TRUE)
d2 <- read.table(files[2], sep=",", header=TRUE)
d <- merge(d1, d2, by="row.names")
colnames(d)[-1] <- c("DNAm", "RNAseq")

d <- data.frame(DNAm=d[,2], RNAseq=d[,3])
gp <- ggplot(d, aes(x=DNAm, y=RNAseq))
gp <- gp + geom_point() + geom_abline(intercept=0, slope=1, col=1) + ggtitle("Concordance `Age` prediction")
gp + theme_bw() + scale_color_npg() + theme(legend.position = "bottom", legend.direction = "horizontal")
ggsave(file.path(path, "Age_con.eps"))


files <- dir(path, pattern="WBCC.*csv", full.names=TRUE)
d1 <- read.table(files[1], sep=",", header=TRUE)
d2 <- read.table(files[2], sep=",", header=TRUE)
d <- merge(d1, d2, by="row.names")

x <- d[,grepl(".*x$", colnames(d))]
y <- d[,grepl(".*y$", colnames(d))]
celltype <- rep(gsub("\\.x$", "", colnames(x)), each=nrow(d))


d <- data.frame(DNAm=as.vector(unlist(x)), RNAseq=as.vector(unlist(y)), celltype=celltype)

dummy <- t(simplify2array(by(d, d$celltype, function(x) range(c(x$DNAm, x$RNAseq)))))
dummy <- data.frame(celltype=rep(rownames(dummy), 2), DNAm=as.vector(dummy), RNAseq=as.vector(dummy))

library(ggplot2)
library(ggsci)
gp <- ggplot(d, aes(x=DNAm, y=RNAseq))
gp <- gp + geom_point() + geom_abline(intercept=0, slope=1, col=2)
gp <- gp + geom_blank(data=dummy)
gp <- gp + facet_wrap(~celltype, scales="free")
gp + theme_bw() + scale_color_npg()
ggsave(file.path(path, "WBCC_con.eps"))

library(ggplot2)
library(ggsci)

d <- data.frame(x = c(rowSums(d1), rowSums(d2)))                
d$type <- c(rep("DNAm", nrow(d1)), rep("RNAseq", nrow(d2)))

gp <- ggplot(d, aes(x=x))
gp <- gp + geom_histogram(colour="black", fill = "white") + xlab("Total Percentage") 
gp + facet_grid(.~type) + theme_bw() + scale_color_npg()
ggsave(file.path(path, "WBCC_total.eps"))

```


## Try out `sparklyr` ##

```{r, eval=FALSE}
##setup spark connection: sc

library(sparklyr)

sc <- spark_connect(master = "local", config = spark_config(), version = "1.6.2")

metadata(bts)$formula <- ~0+Sampling_Age

phenotype <- unlist(get_all_vars(metadata(bts)$formula, data=colData(bts)))
features <- assays(bts)$data

library(dplyr)
data <- t(features[1:1000, 1:100])
data <- cbind(phenotype=phenotype[1:100], data)
rownames(data) <- colnames(features)[1:100]
data[1:5, 1:5]

data_tbl <- copy_to(sc, data)

partitions <- data %>%
  sdf_partition(training = 2/3, test = 1/3, seed = 1099)

fit <- ml_linear_regression(response = "phenotype", features = c("wt", "cyl"))

alpha = 0,
lambda = 0

```


