---
title: "Phenotyping using RNAseq data (IN PROGRESS)"
output:
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
bibliography: BBMRIomics.bib
---

```{r, child="_setup_devel.Rmd"}
```

# Intro #

This document describes how we try to impute some missing phenotypes,
specifically, sex, age, smoking status and white blood cell
composition.

```{r}
preprocess <- function(sumExp) {
    require(SummarizedExperiment)
    require(edgeR)

    y <- DGEList(assays(sumExp)$data)

    message("[preprocess] drop features not expressed")
    ##https://www.bioconductor.org/packages/devel/bioc/vignettes/limma/inst/doc/usersguide.pdf
    A <- rowSums(y$counts)
    isexpr <- A > 50
    y <- y[isexpr,]

    message("[preprocess] normalize")
    y <- calcNormFactors(y) ##normalize takes some time

    message("[preprocess] calculate log counts per million")
    y <- cpm(y, log=TRUE)

    sumExp <- sumExp[isexpr,]
    assays(sumExp)$data <- y ##replace data with filtered subset
    sumExp
}

validate <- function(phenotype, features, train.frac = 2/3, methods = c("ridge", "elastic-net", "pls", "lasso", "gbm", "mboost"), ntop = c(50, 100, 250, 500, 1000, 2500, 5000), nrep = 25, verbose=TRUE) {
    require(omicsPrint)
    require(reshape2)
    require(BiocParallel)
        
    validations <- c()
    for(ntopi in ntop) {
        reps <- bplapply(1:nrep, function(x) phenotyping(phenotype, features, train.frac=train.frac, methods = methods, ntop=ntopi, verbose=verbose)$validation)
        reps <- do.call('rbind', reps)
        validation <- melt(reps, varnames=c("method", "measure"))
        validation$ntop <- ntopi
        validations <- rbind(validations, validation)
        gc()
    }
    validations
}

path <- "/virdir/Scratch/RP3_analysis/imputed_phenotypes"

```

# Inspect missing data #

```{r}
library(BBMRIomics)
library(omicsPrint)
data(rnaSeqData_ReadCounts_BIOS_cleaned)
covariates <- as.data.frame(colData(counts)[, c("Sex", "Sampling_Age", "Smoking", "Lymph_Perc","Mono_Perc", "Eos_Perc", "Neut_Perc")])
##missing covariate summary
missingCov <- apply(covariates, 2, function(x) round(100*table(is.na(x))/length(x),2))
missingCov
##missing sample summary
missingSam <- apply(covariates, 1, function(x) round(100*sum(is.na(x))/length(x),2))
table(missingSam)

tbl <- table(counts$biobank_id, is.na(counts$Sampling_Age))
round(100*tbl[,2]/rowSums(tbl),2)

biobank <- factor(colData(counts)$biobank_id)
missing <- c(4, 82,12,2,2,48)
levels(biobank) <- paste0(levels(biobank), " (", missing, ")")
colData(counts)$biobank_id <- biobank

library(ggplot2)
gp <- ggplot(as.data.frame(colData(counts)), aes(x=Sampling_Age))
gp <- gp + geom_histogram()
gp + facet_wrap(~biobank_id)
ggsave(file.path(path, "Age_RNAseq_distr.eps"))

```

# Predicting Sex #
## Model Selection and Validation ##

First we need to find out which algorithm and which number of features
to use. 

```{r}
data(rnaSeqData_ReadCounts_BIOS_cleaned)
dim(counts)
cnts <- preprocess(counts)
dim(cnts)

metadata(cnts)$formula <- ~0+Sex

phenotype <- unlist(get_all_vars(metadata(cnts)$formula, data=colData(cnts)))
features <- assays(cnts)$data
table(phenotype, useNA="always")

library(BiocParallel)
register(MulticoreParam(10))
validation <- validate(phenotype, features, methods = c("ridge", "elastic-net", "lasso"), ntop = c(5, 10, 25, 50, 100), nrep = 25, verbose=TRUE)

library(ggplot2)
library(ggsci)
gp <- ggplot(subset(validation, measure %in% c("accuracy (Overall)", "f1 (Male)", "f1 (Female)")),
             aes(x=as.factor(ntop), y=value, col=method))
gp <- gp + geom_boxplot()
gp <- gp + facet_wrap(~measure, scales="free", nrow=3, ncol=1)
gp <- gp + xlab("Number of features") + ylab("") + ggtitle("Validation of predictors for `Sex`")
gp + theme_bw() + scale_color_npg() + theme(legend.position = "bottom", legend.direction = "horizontal")

ggsave(file.path(path, "Sex_RNAseq_eval.eps"))
```

## Predict missing ##

```{r}
phenotyped <- phenotyping(phenotype, features, methods = "lasso", ntop=25, verbose=TRUE)
phenotyped$validation
predicted <- phenotyped$predicted
rownames(predicted) <- colnames(features)
write.table(predicted, file=file.path(path, "Sex_RNAseq.csv"), sep=",", quote=FALSE)
```

> TODO add diagnostic plots!

## Inspect selected features ##

It could be interesting to see which genes are the top genes selected
for predicting Age.

> this does not include e.g., the lasso selection

```{r}
library(org.Hs.eg.db)
map <- select(org.Hs.eg.db, keys = names(phenotyped$top), keytype="ENSEMBL", columns=c("SYMBOL", "CHR"))
top <- merge(map, phenotyped$top, by.x="ENSEMBL", by.y="row.names")
top <- top[order(abs(top$y), decreasing=TRUE),]
top

mid <- match(top$ENSEMBL, rownames(features))
y <- features[mid,]
rownames(y) <- top$SYMBOL
rownames(y)[is.na(top$SYMBOL)] <- top$ENSEMBL[is.na(top$SYMBOL)]
rownames(y) <- paste0(rownames(y), " ( chr", top$CHR, ")")
colnames(y) <- NULL

library(ComplexHeatmap)
sex <- colData(cnts)$Sex
sex[is.na(sex)] <- "Unknown"
ha <- HeatmapAnnotation(df = data.frame(Sex = factor(sex)), col = list(Sex = c("Male" =  "blue", "Female" = "pink", "Unknown" = "grey")))

setEPS()
postscript(file.path(path, "Sex_RNAseq_feats.eps"))
Heatmap(y, top_annotation = ha, name="log CPM", row_names_gp = gpar(fontsize = 8))
dev.off()
```


## Detect outliers ##

Outlier or missclassified sample could potentially be wrongly labeled.

Since, we use a subset for training it could be accidentally be
missclassified. Therefore, we run multiple models and collected the
number and frequencies of missclassified samples.


```{r, eval=FALSE}
library(BiocParallel)
register(MulticoreParam(10))
nrep <- 25
predicted <- bplapply(1:nrep, function(x) phenotyping(phenotype, features, methods = "lasso", ntop=25, verbose=FALSE)$predicted)
predicted <- do.call('cbind', predicted)

rownames(predicted) <- names(phenotype) <- colnames(features)

missclassified <- (predicted != phenotype)[!is.na(phenotype), ]
rownames(missclassified) <- colnames(features)[!is.na(phenotype)]
missclassified <- missclassified[rowSums(missclassified) > 0,]
missclass <- 100*rowSums(missclassified)/nrep
missclass[missclass > 50]

x <- cbind(predicted[rownames(predicted) %in% names(missclass[missclass > 50]), 1:5],
      reported=as.character(phenotype[names(phenotype) %in% names(missclass[missclass > 50])])
)
ids <- getView("getIds", usrpwd=RP3_MDB_USRPWD, verbose = FALSE)
subset(ids, uuid %in% rownames(x))

```

# Predicting Age #

## Model selection and Validation ##

First we need to find out which algorithm and which number of features
to use:

```{r}
metadata(cnts)$formula <- ~0+Sampling_Age
phenotype <- unlist(get_all_vars(metadata(cnts)$formula, data=colData(cnts)))
features <- assays(cnts)$data
summary(phenotype)
                       
library(BiocParallel)
register(MulticoreParam(10, log=TRUE))
validation <- validate(phenotype, features, methods = c("ridge", "lasso", "elastic-net", "pls"),
                       ntop = c(100, 250, 500, 1000, 2500, 5000, 10000), nrep = 25, verbose=TRUE)

library(ggplot2)
library(ggsci)
gp <- ggplot(validation, aes(x=as.factor(ntop), y=value, col=method))
gp <- gp + geom_boxplot()
gp <- gp + facet_wrap(~measure, scales="free", nrow=3, ncol=1)
gp <- gp + xlab("Number of features") + ylab("") + ggtitle("Validation of predictors for `Age`")
gp + theme_bw() + scale_color_npg() + theme(legend.position = "bottom", legend.direction = "horizontal")
ggsave(file.path(path, "Age_RNAseq_eval.eps"))
```

some diagnostic plots:

```{r, eval=FALSE}
phenotyped <- phenotyping(phenotype, features, methods = "lasso", ntop=5000, verbose=TRUE)
predicted <- phenotyped$predicted
rownames(predicted) <- colnames(features)

type <- rep("train", length(phenotype))
type[is.na(phenotype)] <- "unknown"
type[phenotyped$testid] <- "test"

phenotype[is.na(phenotype)] <- predicted[is.na(phenotype)]

d <- data.frame(predicted=predicted[,1], measured=phenotype, type=factor(type))

gp <- ggplot(d, aes(x=predicted, y=measured, col=type))
gp <- gp + geom_point() + geom_abline(intercept=0, slope=1, col=1) + ggtitle("Age prediction")
gp + theme_bw() + scale_color_npg() + theme(legend.position = "bottom", legend.direction = "horizontal")
ggsave(file.path(path, "Age_RNAseq_diagn.eps"))


```

## Predict missing ##

```{r}
phenotyped <- phenotyping(phenotype, features, methods = "lasso", ntop=5000, verbose=TRUE)
predicted <- phenotyped$predicted
rownames(predicted) <- colnames(features)
write.table(predicted, file=file.path(path, "Age_RNAseq.csv"), sep=",", quote=FALSE)
```

## Inspect selected features ##


```{r, eval=FALSE}
library(org.Hs.eg.db)
map <- select(org.Hs.eg.db, keys = names(phenotyped$top), keytype="ENSEMBL", columns="SYMBOL")
top <- merge(map, phenotyped$top, by.x="ENSEMBL", by.y="row.names")
head(top[order(abs(top$y), decreasing=TRUE),])

library(irlba)
y <- features[rownames(features) %in% names(phenotyped$top),]
pc <- prcomp_irlba(t(y))
summary(pc)

cols <- rainbow(10)[cut(phenotype, breaks=10)]
plot(pc$x, col=adjustcolor(cols, alpha.f=0.3), pch=16)

```


## Detect outliers ##

```{r, eval=FALSE}
library(BiocParallel)
register(MulticoreParam(10))
nrep <- 100
predicted <- bplapply(1:nrep, function(x) phenotyping(phenotype, features, methods = "lasso", ntop=5000, verbose=FALSE)$predicted)
predicted <- do.call('cbind', predicted)

residuals <- predicted - phenotype

popo <- which(rowSums(abs(residuals) > mean(residuals, na.rm=TRUE) + 3*sd(residuals, na.rm=TRUE)) > 50)

names(residuals) <- colnames(features)

names(residuals)[popo]

library(matrixStats)
mns <- rowMeans(predicted)
sds <- rowSds(predicted)
col <- rep(1, length(mns))
col[popo] <- 2
plot(phenotype, mns, pch=16, cex=sds, col=adjustcolor(col, alpha.f=0.3), bty="n", xlab="reported age", ylab="average predicted age")
abline(0, 1, col=4, lwd=2, lty=2)
grid(col=1)


```


# Predict white blood cell composition #


## Model fitting and model selection ##

```{r, wbccfit}

metadata(cnts)$formula <- ~0 + Lymph_Perc + Mono_Perc + Eos_Perc + Neut_Perc

phenotypes <- get_all_vars(metadata(cnts)$formula, data=colData(cnts))
features <- assays(cnts)$data

complete <- apply(phenotypes, 1, function(x) all(!is.na(x)))
table(complete)

trainId <- sample(which(complete), size = (2/3)*sum(complete))
testId <- setdiff(which(complete), trainId)

length(trainId)
length(testId)

flowcell <- gsub("-.*$", "", cnts$run_id)
batches <- model.matrix(~flowcell)

library(pls)
pls.options(parallel = 5)
predictor <- plsr(log10(phenotypes+1)~batches+data, ncomp = 50,
                  data = list(phenotypes = as.matrix(phenotypes[trainId,]),
                              batches = batches[trainId, ],
                              data=t(features[, trainId])), validation = "CV", keep.model = TRUE)

str(predictor)
summary(predictor)
cumsum(explvar(predictor))
barplot(cumsum(explvar(predictor)), las = 2, ylab = "Cumulative Variance Explained")

validationplot(predictor, val.type = "R2", ncomp = 1:50)
setEPS()
postscript(file.path(path, "WBCC_eval.eps"))
validationplot(predictor, val.type = "RMSEP", ncomp = 1:50)
dev.off()

```


## Evaluate predictions ##

```{r, wbcceval}

predicted <- predict(predictor, newdata = list(data=t(features[, testId]), batches = batches[testId,]) , ncomp = 20)

predicted <- 10^predicted - 1

library(reshape2)
data <- cbind(melt(phenotypes[testId,], value.name="measured"),
              predicted=melt(predicted[,,1])[,3])

by(data, data$variable, function(x) round(cor(x$measured, x$predicted),2))

##hack to get plot area square
dummy <- t(simplify2array(by(data, data$variable, function(x) range(c(x$measured, x$predicted)))))
dummy <- data.frame(variable=rep(rownames(dummy), 2), measured=as.vector(dummy), predicted=as.vector(dummy))

##correlation plot
library(ggplot2)
library(ggsci)
gp <- ggplot(data, aes(x=measured, y=predicted))
gp <- gp + geom_point() + geom_abline(intercept=0, slope=1, col=2)
gp <- gp + geom_blank(data=dummy)
gp <- gp + facet_wrap(~variable, scales="free")
gp + theme_bw() + scale_color_npg()
ggsave(file.path(path, "WBCC_RNAseq_diagn1.eps"))

##bland altman plot
tdata <- data
tdata[,2] <- rowMeans(data[,-1])
tdata[,3] <- data[,2] - data[,3]
colnames(tdata)[2:3] <- c("Average", "Difference")

gp <- ggplot(tdata, aes(x=Average, y=Difference))
gp <- gp + geom_point() + geom_hline(yintercept=0, col=2)
gp <- gp + facet_wrap(~variable, scales="free")
gp + theme_bw() + scale_color_npg()
ggsave(file.path(path, "WBCC_RNAseq_diagn2.eps"))

```



```{r, wbccpred}

predicted <- predict(predictor, newdata = list(data=t(features), batches = batches), ncomp = 20)[,,1]
predicted <- 10^predicted-1

hist(rowSums(predicted), n=50)

dim(predicted)
rownames(predicted) <- colnames(features)
head(predicted)
write.table(predicted, file=file.path(path, "WBCC_RNAseq.csv"),
            sep=",", quote=FALSE)

```

# Predicting Smoking #
## Model Selection and Validation ##

First we need to find out which algorithm and which number of features
to use. 

```{r}
data(rnaSeqData_ReadCounts_BIOS_cleaned)
dim(counts)
cnts <- preprocess(counts)
dim(cnts)

metadata(cnts)$formula <- ~0+Smoking

phenotype <- unlist(get_all_vars(metadata(cnts)$formula, data=colData(cnts)))
features <- assays(cnts)$data

round(100*table(phenotype, useNA="always")/length(phenotype),2)
levels(phenotype) <- c("current smoker", "non-smoker", "non-smoker")
round(100*table(phenotype, useNA="always")/length(phenotype),2)

library(BiocParallel)
register(MulticoreParam(5))
validation <- validate(phenotype, features, methods = c("ridge", "elastic-net", "lasso"), ntop = c(10, 25, 50, 100, 250, 500, 5000, 10000), nrep = 25, verbose=TRUE)

library(ggplot2)
library(ggsci)
gp <- ggplot(subset(validation, measure %in% c("accuracy (Overall)", "f1 (non-smoker)", "f1 (current smoker)")),
             aes(x=as.factor(ntop), y=value, col=method))
gp <- gp + geom_boxplot()
gp <- gp + facet_wrap(~measure, scales="free", nrow=3, ncol=1)
gp <- gp + xlab("Number of features") + ylab("") + ggtitle("Validation of predictors for `Smoking`")
gp + theme_bw() + scale_color_npg() + theme(legend.position = "bottom", legend.direction = "horizontal")

ggsave(file.path(path, "Smoking_RNAseq_eval.eps"))

```

## Predict missing ##

```{r}
phenotyped <- phenotyping(phenotype, features, methods = "elastic-net", ntop=50, verbose=TRUE)
phenotyped$validation
predicted <- phenotyped$predicted
rownames(predicted) <- colnames(features)
write.table(predicted, file=file.path(path, "Smoking_RNAseq.csv"), sep=",", quote=FALSE)
```

> TODO add diagnostic plots!

## Inspect selected features ##

It could be interesting to see which genes are the top genes selected
for predicting Age.

> this does not include e.g., the lasso selection

```{r}
library(org.Hs.eg.db)
map <- select(org.Hs.eg.db, keys = names(phenotyped$top), keytype="ENSEMBL", columns=c("SYMBOL", "CHR"))
top <- merge(map, phenotyped$top, by.x="ENSEMBL", by.y="row.names")
top <- top[order(abs(top$y), decreasing=TRUE),]
top

mid <- match(top$ENSEMBL, rownames(features))
y <- features[mid,]
rownames(y) <- top$SYMBOL
rownames(y)[is.na(top$SYMBOL)] <- top$ENSEMBL[is.na(top$SYMBOL)]
colnames(y) <- NULL

library(ComplexHeatmap)
smoking <- as.character(phenotype)
smoking[is.na(smoking)] <- "Unknown"
ha <- HeatmapAnnotation(df = data.frame(Smoking = factor(smoking)), col = list(Smoking = c("current smoker" =  "blue", "non-smoker" = "pink", "Unknown" = "grey")))

setEPS()
postscript(file.path(path, "Smoking_RNAseq_feats.eps"))
Heatmap(y, top_annotation = ha, name="log CPM", row_names_gp = gpar(fontsize = 8))
dev.off()
```

## Try out `sparklyr` ##

```{r, eval=FALSE}
##setup spark context
library(sparklyr)
sc <- spark_connect(master = "local", config = spark_config(), version = "1.6.2")

metadata(cnts)$formula <- ~0+Sampling_Age
phenotype <- unlist(get_all_vars(metadata(cnts)$formula, data=colData(cnts)))
features <- assays(cnts)$data

library(HDF5Array)
smallMatrix <- DelayedArray(features)
class(smallMatrix)

writeHDF5Dataset(smallMatrix, file="test.hdf5", name="features")

data <- t(features)
data <- cbind(phenotype=phenotype, data)
rownames(data) <- colnames(features)
data[1:5, 1:5]
dim(data)

library(SparkR, lib.loc=".cache/spark/spark-1.6.2-bin-hadoop2.6/R/lib/")
library(rhdf5)
data <- h5read("test.hdf5", name="features")

x <- as.DataFrame(data.frame(data))

write.parquet(data, path="~/test.parquet")

temp_parquet <- tempfile(fileext = ".parquet")

hdf5_to <- function(sc, file) {
  spark_context(sc) %>% 
    invoke("org.apache.spark.ml.Utils.loadHDF5", file)
}

hdf5_to(sc, "test.hdf5")


partitions <- data %>%
  sdf_partition(training = 2/3, test = 1/3, seed = 1099)

fit <- ml_linear_regression(response = "phenotype", features = c("wt", "cyl"))

alpha = 0,
lambda = 0

```
